Live coding environment for a data exploration language

### 2019-02-10 wip-live-programming/paper.tex
Spellchecking

Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some,
can even export the workflow as a script \cite{wrangler}, but data anlysts who prefer working with
code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
Those are text-based, but have a limited model of recomputation. Users structure code in
cells and manually reevaluate cells. Many notebook systems do not track dependencies
between cells, which can lead to inconsistencies; an issue that has been addressed
by e.g.~\citet{dataflow,noworkflow} or \citet{wrattler}.

@@

As an example, the anlysis done by the Financial Times for an article on plastic waste
\cite{ftnotebooks,ftarticle} joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
an extract from one notebook from the data analysis. The code has a number of remarkable properties:
%
\begin{itemize}[itemsep=3pt]
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any top-level functions. Parameterization is done by defining a list of
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options.
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

\item The code relies on external libraries. Our example uses Pandas \cite{pandas},
  which provides opreations for data wrangling such as \ident{merge} to join datasets
  or \ident{drop\_duplicates} to delete rows with duplicate column values. Those libraries are
  external to the data analysis and may even be implemented in other languages -- many Python
  libraries rely on C++.

@@

\begin{figure}[b]
  \centering
  \subcaptionbox{
    \raggedright\footnotesize The Gamma script to aggregate Olympic medals. The analyst selects
      Rio 2016 games and counts the number of distinct events per athlete. After typing `.',
      a type provider offers further aggregation operations.
    \label{fig:gamma-dot}
  }{\includegraphics[scale=0.26]{gamma1.png}}
  \subcaptionbox{
    \raggedright\footnotesize Live preview in The Gamma programming environment developed based on
      the theory presented in this paper. The table is produced as
      the data anlyst types and updates on-the-fly to show result at the current cursor position.
    \label{fig:gamma-preview}
  }{\includegraphics[scale=0.26]{gamma2.png}}
  \caption{\small Data exploration in The Gamma -- using a type provider (left) and with a live
    preview (right).}
  \label{fig:gamma-intro}
\end{figure}

@@

As illustrated by \citet{gamma}, the combination of type providers and auto-complete on~`.' makes it
possible to complete a large number of data exploration tasks through a very simple interaction of
selecting operations from a list. An example in Figure~\ref{fig:gamma-dot} summarizes data on
Olympic medals using a type provider. The names of identifiers such as \qident{sum Bronze} do not have
any special meaning. They are merely members generated by the type provider, based on information
about the data source. The type provider used in this example generates an object with members for
different data transformations, such as \qident{group data}, which return furhter objects with
members for specifying transformation parameters, such as selecting the grouping key using
\qident{by Athlete}.

The Gamma language is richer, but the example in Figure~\ref{fig:gamma-dot} shows that non-trivial
data exploration can be done using a very simple language. The assumptions about strucutre of
code that are explicit in The Gamma are implicitly present in Python and R data analyses
produced by journalists, economists and other users with other than programming background.
When we refer to The Gamma in this paper, readers not familiar with it can assume that we focus
on a subset of Python.

@@

An example of a live preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
Gamma program consits of a list of commands which are either expressions or let bindings. Our
editor displays a live preview below the command that the user is currently editing. The preview
shows the result of evaluating the expression or the value assigned to a bound variable. When the
user changes the code, the preview is updated automatically, without any additional interaction
with the user.

There is a number of guiding principles that our work follows. First, we allow the analyst
to edit code in an unrestricted form in a text editor. Although structured editors provide an
appearling alternative and make recomputation easier, we prefer the transparency
and openness of plain text. Second, we focus on the scenario when code changes, but input
does not. Rapid feedback allows the analyst to quickly adapt code to correctly handle
corner cases that typical analysis involves. In contrast to work on incremental computation,
we do not consider the case when data changes, although supporting interactive data exploration
of streaming data is an interesting future direction.

@@

The \emph{data exploration calculus} is a small, formally tractable language for data exploration.
The calculus is not Turing-complete and it can only be used together with external libraries that
define what objects are available and what the behaviour of their members is. This is
sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
define the caluclus in this section and then use it as the basis for discussing our live preview
mechanism in Section~\ref{sec:formal}.

@@

The data exploration calculus is a call-by-value language. We model the evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
objects $o_1;\; \ldots;\; o_n$ (one object for each command, including let bindings) which can be
displayed as intermediate results of the data analysis. The operational semantics is parameterized
by a relation $\rightsquigarrow_\epsilon$ that models the functionality of the external libraries
used with the calculus and defines reduction rules for member accesses. The relation has the
following form:
%
\begin{equation*}
o_1.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o_2
\end{equation*}
%
The operation is invoked on an object and takes values (objects or function values) as arguments.
It always results in an opaque object.
Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
$C_c$ specifies evluation of a command and $C_p$ defines top-to-bottom evaluation of a program.
The rule \rname{external} applies reductions provided by an external library in a
call-by-value order and \rname{let} substitutes a value of an evaluated variable in
all subsequent commands. Note that \rname{let} leaves the value of the variable
in the resulting list of commands.

@@

\begin{definition}[External library]
\label{def:external}
An external library consists of a set of objects $O$ and a reduction relation
$\rightsquigarrow_\epsilon$ that satsifies the following two properties:
\begin{itemize}
\item[-] \emph{Completeness.} For all $o, m, i$ and all $v_1, \ldots, v_i$, there exists $o'$ such
  that $o.m(v_1, \ldots, v_i) \rightsquigarrow_\epsilon o'$.
\vspace{0.5em}
\item[-] \emph{Compositionality.} For observationally equivalent arguments, the reduction returns
  the same object, i.e.~given $e_0, e_1, \ldots, e_n$ and $e'_0, e'_1, \ldots, e'_n$ and $m$ such that
  $e_0.m(e_1, \ldots, e_n) \rightsquigarrow^{*} o$ and $e'_0.m(e'_1, \ldots, e'_n) \rightsquigarrow^{*} o'$ then
  if for any contexts $C_0, C_1, \ldots, C_n$ it holds that if $C_i[e_i] \rightsquigarrow^{*} o_i$ and
  $C_i[e'_i] \rightsquigarrow^{*} o_i$ for some $o_i$ then $o = o'$.
\end{itemize}
\end{definition}

@@

Furthermore, the graph edges are labeled with labels indicating the kind of dependency. For
a method call, the labels are ``first argument'', ``second argument'' and so on. Writing
$s$ for symbols and $i$ for integers, nodes (vertices) $\bndclr{v}$ and edge labels $\blblclr{l}$
are defined as:
%
\begin{equation*}
\begin{array}{rcll}
\bndclr{v}&=&\bnd{val}(o) \lsep \bnd{var}(x)\lsep \bnd{mem}(m, s)\lsep \bnd{fun}(x, s)&(\textit{Vertices})\\
\blblclr{l}&=&\blbl{body} \lsep \blbl{arg}(i)&(\textit{Edge labels})\\
\end{array}
\end{equation*}
%
The \bnd{val} node represents a primitive value and contains the object itself. Two occurrences
of $\num{10}$ in the source code will be represented by the same node. Member access \bnd{mem}
contains the member name, together with a unique symbol $s$ -- two member access nodes with
different dependencies will contain a different symbol. Dependencies of member access are labeled
with \blbl{arg} indicating the index of the argument (the instance has index $0$ and arguments
start with $1$). Finally, nodes \bnd{fun} and \bnd{var} represent function values and variables
bound by $\lambda$ abstraction. For simplicity, we use variable names rather than de Bruijn
indices and so renaming a bound variable forces recomputation.

@@

\subsubsection{Reusing graph nodes}
The binding process takes an expression and constructs a dependency graph reusing existing nodes
when possible. For this, we use a lookup table of member access and function value nodes. The key
in the lookup table is formed by a node kind (for disambiguation) together with a list of
dependencies. A node kind is a member access containing the member name or a function containing
the bound variable name; a lookup table $\Delta$ then maps a node kind with a list of dependencies
to a cached node:
%
\begin{equation*}
\begin{array}{ll}
\bkndclr{k}~=~\bknd{fun}(x)~|~\bknd{mem}(m)&(\textit{Node kinds})\\[0.2em]
\Delta(\bkndclr{k}, [(\bndclr{v_1},\blblclr{l_1}), \ldots, (\bndclr{v_n},\blblclr{l_n})])  & (\textit{Lookup for a node})
\end{array}
\end{equation*}
%
The example on the second line looks for a node of a kind $\bkndclr{k}$ that has dependencies
$\bndclr{v_1}, \ldots, \bndclr{v_n}$ labeled with labels $\blblclr{l_1}, \ldots, \blblclr{l_n}$.
We write $\Delta(\bndclr{k}, l)\!\downarrow$ when a value for a given key is not defined.
For example, when creating the graph in Figure~\ref{fig:dep-graph-b},
we perform the following lookup for the \ident{skip} member access:
%
\[ \Delta(\bknd{mem}(\ident{skip}), [(\bnd{var}(\ident{data}),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) \]
%
The lookup returns the node $\bnd{mem}(\ident{skip}, s_0)$ known from the previous step. We then perform
the following lookup for the \ident{take} member access:
%
\[ \Delta(\bknd{mem}(\ident{take}), [(\bnd{mem}(\ident{skip}, s_0),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) \]
%
In the previous graph, the argument of \ident{take} was $\num{15}$ rather than $\num{10}$ and so
this lookup fails. We then construct a new node $\bnd{mem}(\ident{take}, s_2)$ using a fresh
symbol $s_2$.

@@

When applied on an expression $e$, binding $\ident{bind-expr}_{\Gamma,\Delta}(e)$ returns
a node $\bndclr{v}$ corresponding to the expression $e$ paired with a dependency graph $(V, E)$.
In the graph, $V$ is a set of nodes $\bndclr{v}$ and $E$ is a set of labeled edges
$(\bndclr{v_1}, \bndclr{v_2}, \blblclr{l})$. We attach the label directly to the edge rather than
keeping a separate colouring function as this makes the formalisation simpler.
The $\ident{bind-prog}_{\Gamma,\Delta}$ function works similarly, but takes a sequence of
commands and returns a sequence of nodes.

\subsubsection{Binding expressions.} When binding member access, we use \ident{bind-expr} recursively
to get a node and a dependency graph for each sub-expression. The nodes representing sub-expressions are
then used as dependencies for lookup into $\Delta$, together with their labels. If a node already
exists in $\Delta$ it is reused (1). Alternativelym, we create a new node containing a fresh symbol~(2).

@@

Programs in the data exploration calculus consist of sequence of commands. Those can always be
evaluated to a value with a preview that can be displayed to the user. However, we also support
previews for all sub-expressions. This can be problematic if the current sub-expression is inside
the body of a function. For example:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{top} = \ident{movies}.\ident{take}(\num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased}().\ident{format}(\str{dd-mm-yyyy}))
\end{array}
\end{equation*}
%
Here, we can directly evaluate sub-expressions \ident{movies} and $\ident{movies}.\ident{take}(\num{10})$,
but not $x.\ident{getReleased}()$ and $x.\ident{getReleased}().\ident{format}(\str{dd-mm-yyyy})$
because they contain a free variable $x$. Our preview evaluation algorithm addresses this by
producing two kinds of previews. A \emph{fully evaluated preview} is just a value, while
a \emph{delayed preview} is a partially evalauted expression with free variables:

@@

\subsection{Evaluation of previews}
The evaluation of previews is defined in Figure~\ref{fig:eval}. Given a dependency graph $(V, E)$,
we define a relation $\bndclr{v}\Downarrow d$ that evaluates a sub-expression corresponding to
the node $\bndclr{v}$ to a possibly delayed preview $d$. The nodes $V$ and edges $E$ of the
dependency graph are parameters of the $\Downarrow$ relation, but they do not change during the
evaluation and so we do not explicity write them.

@@

The semantics of delayed previews can be more formally captured by Contextual Modal Type Theory
(CMTT) \cite{cmtt} and comonads \cite{cmtt-denotation}. In CMTT, $\lbrack \Psi \rbrack A$ denotes
that a proposition $A$ is valid in context $\Psi$, which is similar to our delayed previews written
as $\llbracket A \rrbracket_\Psi$. CMTT defines rules for composing context-dependent propositions
that would allow us to express the splicing operation used in \rname{mem-expr}. In categorical
terms, the context-dependent proposition can be modeled as an indexed comonad \cite{effectrev,graded}.
The evaluation of a preview with no context dependencies (built implicitly into our evaluation rules)
corresponds to the counit operation of a comonad and would be explicitly written as
$\llbracket A \rrbracket_\emptyset \rightarrow A$.

@@

The of preview-preserving code edit operations includes several ways of introducing a let
binding, several ways of eliminating a let binding and two edit operations where the data anlyst
modifies an unrelated part of the program. It is worth noting that the list is not exhaustive.
Rather, it aims to illustrate some of the typical operations that the data analyst might perform
when writing code.

@@

For the purpose of the evaluation, we use a simple image manipulation library that provides
operations for loading, greyscaling, bluring and combining images. We compare delays in
updating the preview for three different evaluation strategies, while performing the same
sequence of code edit operations. Using image processing as an example gives us a way to
visualize the reuse of previously computed values. As in a typical data exploration scenario,
the individual operations are relatively expensive compared to the overheads of building the
dependency graph.

@@

\begin{enumerate}[(a)]
\item Loading image for the first time incurs small extra overhead in the live strategy.
\item Greyscaling using the live strategy does not need to re-load the image.
\item Accessing the blur member without arguments causes delay for call-by-value strategy.
\item When varying the parameter of blur, the live strategy reuses previously greyscaled image.
\item Introducing let binding does not cause recomputation when using live strategy.
\item As in (c), accessing combine member without valud argument only affects call-by-value.
\item Varying parameter of combine in live strategy does not recompute blur and is much faster.
\item Introducing let binding, again, causes full recomputation for lazy can call-by-value.
\end{enumerate}

@@

In Section~\ref{sec:background-jupyter}, we motivated our work by considering how journalists
explore open data. In addition to the theoretical and experimental work presented in this paper,
we also implemented an online data exploration environment, equipped with live editor for The
Gamma language that provides instananeous feedback during coding. The environment uses the
principles presented in this paper to build a more comprehensive system that allows users,
such as journalists, to analyse, summarize and visualize open data. In this section, we briefly
report on our experience with the system.

@@

\begin{itemize}[itemsep=3pt]
\item The left screenshot uses a type provider for data aggregation introduced by \citet{gamma}
  and briefly discussed in Secton~\ref{sec:background-gamma}. For the purpose of
  this paper, the type provider is an external library (with objects, members and reduction relation).
  Using the type provider relies on editor auto-complete which is based on type information. To support
  this, we implement type checking, as discussed in Section~\ref{sec:extra-types}, reusing the mechanism
  of live preview evaluation.

\item The live preview for code written using the data agregation type provider (left) is shown
  in a panel below the current command. This shows the preview for a sub-expression based on the
  position of the cursor (i.e.~raw data before grouping in the screenshot). In addition to the
  preview, our environment also generates tabs that let the user navigate to individual steps of
  the data transformation for code using the data aggregation type provider.

@@

\noindent
Conducting an extensive user study that would evaluate the usability of the system from a
human-computer ineraction perspective is left for future work. To give anecdotal evidence of
the usability, the example shown in Figure~\ref{fig:gallery} is an analysis developed by an
attendee of our tutorial at Mozilla Festival 2017 who had no prior programming experience.

@@

\subsubsection{Type checking}
The structure of typing rules is similar to the evaluation relation $\bndclr{v}\Downarrow d$
defined earlyer. Given a dependency graph $(V, E)$, we define typing judgements in the form
$\bndclr{v}\vdash \tau$. The type $\tau$ can be a primitive type, a function $\tau \rightarrow \tau$
or an object type $\{m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n\}$ with member types
$\sigma = (\tau_1, \ldots, \tau_n) \rightarrow \tau$.

@@

The results of type checking can be cached and reused in the same way as live previews, although
we leave out the details. A property aking to correctness (Theorem~\ref{thm:correcntess})
requires defining standard type checking over the structure of expressions, which we also
omit for space reasons.

@@

Providing live previews inside ordinary functions is problematic, because we do not have readily
available values for input parameters and our mechanism for lambda functions only provides
delayed previews inside body of a function. We believe that extending the data exploration calculus
with an abstraction mechanism that would support development with instantaneous feedaback is an
interesting design problem and we briefly outline a possible solution here.

@@

\noindent
We introduce two new kinds of commands: a labelled sequence of commands and a referenceto a label.
When evaluating, the command $\textit{lbl}$ is replaced with the associated sequence of
commands $p$ before any other reductions. Consequently, variables used in the labelled block are
dynamically scoped and we can use let binding to redefine a value of a variable before invoking
the block repeatedly. The correct use of dynamic scoping can be checked using
coeffects \cite{coeffects}.

@@

\section{Appenidx - Type checking}
\label{sec:app-types}

>>>>

Many visual data exploration tools support interactivity \cite{control,tableau,vizdom} and some,
can even export the workflow as a script \cite{wrangler}, but data analysts who prefer working with
code typically resort to notebook systems such as Jupyter or R Markdown \cite{jupyter,rmarkdown}.
Those are text-based, but have a limited model of recomputation. Users structure code in
cells and manually reevaluate cells. Many notebook systems do not track dependencies
between cells, which can lead to inconsistencies; an issue that has been addressed
by e.g.~\citet{dataflow,noworkflow} or \citet{wrattler}.

@@

As an example, the analysis done by the Financial Times for an article on plastic waste
\cite{ftnotebooks,ftarticle} joins datasets from Eurostat, UN Comtrade and more, aggregates the data and
builds a visualization comparing waste flows in 2017 and 2018. Figure~\ref{fig:ft-uncomtrade} shows
an extract from one notebook from the data analysis. The code has a number of remarkable properties:
%
\begin{itemize}[itemsep=3pt]
\item There is no abstraction. The analysis uses lambda functions as arguments to library calls,
  but it does not define any top-level functions. Parameterization is done by defining a list of
  inputs and then having a for loop, or by defining a top-level variable and setting its value.
  In our example, \ident{material} is set to \str{plastics} and a comment lists other options.
  This lets the analyst easily see and check the results of intermediate steps of the analysis.

\item The code relies on external libraries. Our example uses Pandas \cite{pandas},
  which provides operations for data wrangling such as \ident{merge} to join datasets
  or \ident{drop\_duplicates} to delete rows with duplicate column values. Those libraries are
  external to the data analysis and may even be implemented in other languages -- many Python
  libraries rely on C++.

@@

\begin{figure}[b]
  \centering
  \subcaptionbox{
    \raggedright\footnotesize The Gamma script to aggregate Olympic medals. The analyst selects
      Rio 2016 games and counts the number of distinct events per athlete. After typing `.',
      a type provider offers further aggregation operations.
    \label{fig:gamma-dot}
  }{\includegraphics[scale=0.26]{gamma1.png}}
  \subcaptionbox{
    \raggedright\footnotesize Live preview in The Gamma programming environment developed based on
      the theory presented in this paper. The table is produced as
      the data analyst types and updates on-the-fly to show result at the current cursor position.
    \label{fig:gamma-preview}
  }{\includegraphics[scale=0.26]{gamma2.png}}
  \caption{\small Data exploration in The Gamma -- using a type provider (left) and with a live
    preview (right).}
  \label{fig:gamma-intro}
\end{figure}

@@

As illustrated by \citet{gamma}, the combination of type providers and auto-complete on~`.' makes it
possible to complete a large number of data exploration tasks through a very simple interaction of
selecting operations from a list. An example in Figure~\ref{fig:gamma-dot} summarizes data on
Olympic medals using a type provider. The names of identifiers such as \qident{sum Bronze} do not have
any special meaning. They are merely members generated by the type provider, based on information
about the data source. The type provider used in this example generates an object with members for
different data transformations, such as \qident{group data}, which return further objects with
members for specifying transformation parameters, such as selecting the grouping key using
\qident{by Athlete}.

The Gamma language is richer, but the example in Figure~\ref{fig:gamma-dot} shows that non-trivial
data exploration can be done using a very simple language. The assumptions about structure of
code that are explicit in The Gamma are implicitly present in Python and R data analyses
produced by journalists, economists and other users with other than programming background.
When we refer to The Gamma in this paper, readers not familiar with it can assume that we focus
on a subset of Python.

@@

An example of a live preview is shown in Figure~\ref{fig:gamma-preview}. As noted earlier, The
Gamma program consist of a list of commands which are either expressions or let bindings. Our
editor displays a live preview below the command that the user is currently editing. The preview
shows the result of evaluating the expression or the value assigned to a bound variable. When the
user changes the code, the preview is updated automatically, without any additional interaction
with the user.

There is a number of guiding principles that our work follows. First, we allow the analyst
to edit code in an unrestricted form in a text editor. Although structured editors provide an
appealing alternative and make recomputation easier, we prefer the transparency
and openness of plain text. Second, we focus on the scenario when code changes, but input
does not. Rapid feedback allows the analyst to quickly adapt code to correctly handle
corner cases that typical analysis involves. In contrast to work on incremental computation,
we do not consider the case when data changes, although supporting interactive data exploration
of streaming data is an interesting future direction.

@@

The \emph{data exploration calculus} is a small, formally tractable language for data exploration.
The calculus is not Turing-complete and it can only be used together with external libraries that
define what objects are available and what the behaviour of their members is. This is
sufficient to capture the simple data analyses discussed in Section~\ref{sec:background}. We
define the calculus in this section and then use it as the basis for discussing our live preview
mechanism in Section~\ref{sec:formal}.

@@

The data exploration calculus is a call-by-value language. We model the evaluation as a small-step
reduction $\rightsquigarrow$. Fully evaluating a program results in an irreducible sequence of
objects $o_1;\; \ldots;\; o_n$ (one object for each command, including let bindings) which can be
displayed as intermediate results of the data analysis. The operational semantics is parameterized
by a relation $\rightsquigarrow_\epsilon$ that models the functionality of the external libraries
used with the calculus and defines reduction rules for member accesses. The relation has the
following form:
%
\begin{equation*}
o_1.m(v_1, \ldots, v_n) \rightsquigarrow_\epsilon o_2
\end{equation*}
%
The operation is invoked on an object and takes values (objects or function values) as arguments.
It always results in an opaque object.
Figure~\ref{fig:dec-calculus} defines the reduction rules in terms of $\rightsquigarrow_\epsilon$
and evaluation contexts; $C_e$ specifies left-to-right evaluation of arguments of a method call,
$C_c$ specifies evaluation of a command and $C_p$ defines top-to-bottom evaluation of a program.
The rule \rname{external} applies reductions provided by an external library in a
call-by-value order and \rname{let} substitutes a value of an evaluated variable in
all subsequent commands. Note that \rname{let} leaves the value of the variable
in the resulting list of commands.

@@

\begin{definition}[External library]
\label{def:external}
An external library consists of a set of objects $O$ and a reduction relation
$\rightsquigarrow_\epsilon$ that satisfies the following two properties:
\begin{itemize}
\item[-] \emph{Completeness.} For all $o, m, i$ and all $v_1, \ldots, v_i$, there exists $o'$ such
  that $o.m(v_1, \ldots, v_i) \rightsquigarrow_\epsilon o'$.
\vspace{0.5em}
\item[-] \emph{Compositionality.} For observationally equivalent arguments, the reduction returns
  the same object, i.e.~given $e_0, e_1, \ldots, e_n$ and $e'_0, e'_1, \ldots, e'_n$ and $m$ such that
  $e_0.m(e_1, \ldots, e_n) \rightsquigarrow^{*} o$ and $e'_0.m(e'_1, \ldots, e'_n) \rightsquigarrow^{*} o'$ then
  if for any contexts $C_0, C_1, \ldots, C_n$ it holds that if $C_i[e_i] \rightsquigarrow^{*} o_i$ and
  $C_i[e'_i] \rightsquigarrow^{*} o_i$ for some $o_i$ then $o = o'$.
\end{itemize}
\end{definition}

@@

Furthermore, the graph edges are labelled with labels indicating the kind of dependency. For
a method call, the labels are ``first argument'', ``second argument'' and so on. Writing
$s$ for symbols and $i$ for integers, nodes (vertices) $\bndclr{v}$ and edge labels $\blblclr{l}$
are defined as:
%
\begin{equation*}
\begin{array}{rcll}
\bndclr{v}&=&\bnd{val}(o) \lsep \bnd{var}(x)\lsep \bnd{mem}(m, s)\lsep \bnd{fun}(x, s)&(\textit{Vertices})\\
\blblclr{l}&=&\blbl{body} \lsep \blbl{arg}(i)&(\textit{Edge labels})\\
\end{array}
\end{equation*}
%
The \bnd{val} node represents a primitive value and contains the object itself. Two occurrences
of $\num{10}$ in the source code will be represented by the same node. Member access \bnd{mem}
contains the member name, together with a unique symbol $s$ -- two member access nodes with
different dependencies will contain a different symbol. Dependencies of member access are labelled
with \blbl{arg} indicating the index of the argument (the instance has index $0$ and arguments
start with $1$). Finally, nodes \bnd{fun} and \bnd{var} represent function values and variables
bound by $\lambda$ abstraction. For simplicity, we use variable names rather than de Bruijn
indices and so renaming a bound variable forces recomputation.

@@

\subsubsection{Reusing graph nodes}
The binding process takes an expression and constructs a dependency graph reusing existing nodes
when possible. For this, we use a lookup table of member access and function value nodes. The key
in the lookup table is formed by a node kind (for disambiguation) together with a list of
dependencies. A node kind is a member access containing the member name or a function containing
the bound variable name; a lookup table $\Delta$ then maps a node kind with a list of dependencies
to a cached node:
%
\begin{equation*}
\begin{array}{ll}
\bkndclr{k}~=~\bknd{fun}(x)~|~\bknd{mem}(m)&(\textit{Node kinds})\\[0.2em]
\Delta(\bkndclr{k}, [(\bndclr{v_1},\blblclr{l_1}), \ldots, (\bndclr{v_n},\blblclr{l_n})])  & (\textit{Lookup for a node})
\end{array}
\end{equation*}
%
The example on the second line looks for a node of a kind $\bkndclr{k}$ that has dependencies
$\bndclr{v_1}, \ldots, \bndclr{v_n}$ labelled with labels $\blblclr{l_1}, \ldots, \blblclr{l_n}$.
We write $\Delta(\bndclr{k}, l)\!\downarrow$ when a value for a given key is not defined.
For example, when creating the graph in Figure~\ref{fig:dep-graph-b},
we perform the following lookup for the \ident{skip} member access:
%
\[ \Delta(\bknd{mem}(\ident{skip}), [(\bnd{var}(\ident{data}),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) \]
%
The lookup returns the node $\bnd{mem}(\ident{skip}, s_0)$ known from the previous step. We then perform
the following lookup for the \ident{take} member access:
%
\[ \Delta(\bknd{mem}(\ident{take}), [(\bnd{mem}(\ident{skip}, s_0),\blbl{arg}(0)), (\bnd{val}(\num{10}), \blbl{arg}(1))]) \]
%
In the previous graph, the argument of \ident{take} was $\num{15}$ rather than $\num{10}$ and so
this lookup fails. We then construct a new node $\bnd{mem}(\ident{take}, s_2)$ using a fresh
symbol $s_2$.

@@

When applied on an expression $e$, binding $\ident{bind-expr}_{\Gamma,\Delta}(e)$ returns
a node $\bndclr{v}$ corresponding to the expression $e$ paired with a dependency graph $(V, E)$.
In the graph, $V$ is a set of nodes $\bndclr{v}$ and $E$ is a set of labelled edges
$(\bndclr{v_1}, \bndclr{v_2}, \blblclr{l})$. We attach the label directly to the edge rather than
keeping a separate colouring function as this makes the formalisation simpler.
The $\ident{bind-prog}_{\Gamma,\Delta}$ function works similarly, but takes a sequence of
commands and returns a sequence of nodes.

\subsubsection{Binding expressions.} When binding member access, we use \ident{bind-expr} recursively
to get a node and a dependency graph for each sub-expression. The nodes representing sub-expressions are
then used as dependencies for lookup into $\Delta$, together with their labels. If a node already
exists in $\Delta$ it is reused (1). Alternatively, we create a new node containing a fresh symbol~(2).

@@

Programs in the data exploration calculus consist of sequence of commands. Those can always be
evaluated to a value with a preview that can be displayed to the user. However, we also support
previews for all sub-expressions. This can be problematic if the current sub-expression is inside
the body of a function. For example:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{top} = \ident{movies}.\ident{take}(\num{10}).\ident{map}(\lambda x \rightarrow x.\ident{getReleased}().\ident{format}(\str{dd-mm-yyyy}))
\end{array}
\end{equation*}
%
Here, we can directly evaluate sub-expressions \ident{movies} and $\ident{movies}.\ident{take}(\num{10})$,
but not $x.\ident{getReleased}()$ and $x.\ident{getReleased}().\ident{format}(\str{dd-mm-yyyy})$
because they contain a free variable $x$. Our preview evaluation algorithm addresses this by
producing two kinds of previews. A \emph{fully evaluated preview} is just a value, while
a \emph{delayed preview} is a partially evaluated expression with free variables:

@@

\subsection{Evaluation of previews}
The evaluation of previews is defined in Figure~\ref{fig:eval}. Given a dependency graph $(V, E)$,
we define a relation $\bndclr{v}\Downarrow d$ that evaluates a sub-expression corresponding to
the node $\bndclr{v}$ to a possibly delayed preview $d$. The nodes $V$ and edges $E$ of the
dependency graph are parameters of the $\Downarrow$ relation, but they do not change during the
evaluation and so we do not explicitly write them.

@@

The semantics of delayed previews can be more formally captured by Contextual Modal Type Theory
(CMTT) \cite{cmtt} and comonads \cite{cmtt-denotation}. In CMTT, $\lbrack \Psi \rbrack A$ denotes
that a proposition $A$ is valid in context $\Psi$, which is similar to our delayed previews written
as $\llbracket A \rrbracket_\Psi$. CMTT defines rules for composing context-dependent propositions
that would allow us to express the splicing operation used in \rname{mem-expr}. In categorical
terms, the context-dependent proposition can be modelled as an indexed comonad \cite{effectrev,graded}.
The evaluation of a preview with no context dependencies (built implicitly into our evaluation rules)
corresponds to the counit operation of a comonad and would be explicitly written as
$\llbracket A \rrbracket_\emptyset \rightarrow A$.

@@

The of preview-preserving code edit operations includes several ways of introducing a let
binding, several ways of eliminating a let binding and two edit operations where the data analyst
modifies an unrelated part of the program. It is worth noting that the list is not exhaustive.
Rather, it aims to illustrate some of the typical operations that the data analyst might perform
when writing code.

@@

For the purpose of the evaluation, we use a simple image manipulation library that provides
operations for loading, greyscaling, blurring and combining images. We compare delays in
updating the preview for three different evaluation strategies, while performing the same
sequence of code edit operations. Using image processing as an example gives us a way to
visualize the reuse of previously computed values. As in a typical data exploration scenario,
the individual operations are relatively expensive compared to the overheads of building the
dependency graph.

@@

\begin{enumerate}[(a)]
\item Loading image for the first time incurs small extra overhead in the live strategy.
\item Greyscaling using the live strategy does not need to re-load the image.
\item Accessing the blur member without arguments causes delay for call-by-value strategy.
\item When varying the parameter of blur, the live strategy reuses previously greyscaled image.
\item Introducing let binding does not cause recomputation when using live strategy.
\item As in (c), accessing combine member without valid argument only affects call-by-value.
\item Varying parameter of combine in live strategy does not recompute blur and is much faster.
\item Introducing let binding, again, causes full recomputation for lazy can call-by-value.
\end{enumerate}

@@

In Section~\ref{sec:background-jupyter}, we motivated our work by considering how journalists
explore open data. In addition to the theoretical and experimental work presented in this paper,
we also implemented an online data exploration environment, equipped with live editor for The
Gamma language that provides instantaneous feedback during coding. The environment uses the
principles presented in this paper to build a more comprehensive system that allows users,
such as journalists, to analyse, summarize and visualize open data. In this section, we briefly
report on our experience with the system.

@@

\begin{itemize}[itemsep=3pt]
\item The left screenshot uses a type provider for data aggregation introduced by \citet{gamma}
  and briefly discussed in Section~\ref{sec:background-gamma}. For the purpose of
  this paper, the type provider is an external library (with objects, members and reduction relation).
  Using the type provider relies on editor auto-complete which is based on type information. To support
  this, we implement type checking, as discussed in Section~\ref{sec:extra-types}, reusing the mechanism
  of live preview evaluation.

\item The live preview for code written using the data aggregation type provider (left) is shown
  in a panel below the current command. This shows the preview for a sub-expression based on the
  position of the cursor (i.e.~raw data before grouping in the screenshot). In addition to the
  preview, our environment also generates tabs that let the user navigate to individual steps of
  the data transformation for code using the data aggregation type provider.

@@

\noindent
Conducting an extensive user study that would evaluate the usability of the system from a
human-computer interaction perspective is left for future work. To give anecdotal evidence of
the usability, the example shown in Figure~\ref{fig:gallery} is an analysis developed by an
attendee of our tutorial at Mozilla Festival 2017 who had no prior programming experience.

@@

\subsubsection{Type checking}
The structure of typing rules is similar to the evaluation relation $\bndclr{v}\Downarrow d$
defined earlier. Given a dependency graph $(V, E)$, we define typing judgements in the form
$\bndclr{v}\vdash \tau$. The type $\tau$ can be a primitive type, a function $\tau \rightarrow \tau$
or an object type $\{m_1\!:\!\sigma_1, \ldots, m_n\!:\!\sigma_n\}$ with member types
$\sigma = (\tau_1, \ldots, \tau_n) \rightarrow \tau$.

@@

The results of type checking can be cached and reused in the same way as live previews, although
we leave out the details. A property akin to correctness (Theorem~\ref{thm:correcntess})
requires defining standard type checking over the structure of expressions, which we also
omit for space reasons.

@@

Providing live previews inside ordinary functions is problematic, because we do not have readily
available values for input parameters and our mechanism for lambda functions only provides
delayed previews inside body of a function. We believe that extending the data exploration calculus
with an abstraction mechanism that would support development with instantaneous feedback is an
interesting design problem and we briefly outline a possible solution here.

@@

\noindent
We introduce two new kinds of commands: a labelled sequence of commands and a reference to a label.
When evaluating, the command $\textit{lbl}$ is replaced with the associated sequence of
commands $p$ before any other reductions. Consequently, variables used in the labelled block are
dynamically scoped and we can use let binding to redefine a value of a variable before invoking
the block repeatedly. The correct use of dynamic scoping can be checked using
coeffects \cite{coeffects}.

@@

\section{Appendix - Type checking}
\label{sec:app-types}
